---
title: "LML Session 6: Tidy data"
author: "Andres Aravena"
date: "21/11/2018"
output:
  ioslides_presentation:
    logo: ../../../images/dry-lab.svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev.args=list(bg="transparent"))
knitr::opts_knit$set(rooo.dir="~/Web/blog/_slides/2018/dry-lab")
```

## Data tidying
One of the important cleaning processes in the practice of data science

*Tidy data sets* have structure and working with them is easy

They are easy to manipulate, model and visualize

Tidy data sets are arranged such that each variable is a column and each observation (or case) is a row

## Characteristics

+ Each variable you measure should be in one column.
+ Each different observation of that variable should be in a different row.
+ There should be one table for each "kind" of variable.
+ If you have multiple tables, they should include a column in the table that allows them to be linked.

::: source
+ Jeff Leek, "The Elements of Data Analytic Style" (2015)
+ Wikipedia
:::

## Onur's data {.no-gap}
```{r}
heroes <- read.csv("heroes.txt", row.names=1, stringsAsFactors=FALSE)
plot(Int ~ Agi, data=heroes, col="black", bg=as.factor(Main), pch=21)
```

## What is the question?

We know the classes for this sample

We can ask two questions

1. What is the class of a new observation?

2. What is the rule that defines each class?

## Data for Training
We use our data to *train* a machine learning method

Typical data is a numeric matrix and a vector of *labels*

For each individual we have several *features*. This is one row in the matrix.

We also have a *label* for each individual. This is how we teach the system.

## Fitting and over-fitting
*Fitting* is the process that the machine does to learn the relationship between *features* and *labels*

We want to find a general rule from a few examples

The problem is that sometimes we learn too much about the sample, and we cannot generalize

When that happens, we say that the model is _over-fitted_

## How to avoid over-fitting
We need to measure the quality of the learning using **different** data

We have to separate our samples in two groups:

+ A training set
+ A test set

That means that we do not use all the data for training. We need even more data

## Extreme case: Leave-one-out

+ We choose one sample as "test".
    + We use the rest for training. 
+ Then we test with the single sample.
+ If the predicted label is correct, we are fine

Then we repeat all the process.

The quality of the method depends on the rate-of-success

## A first method: K nearest neighbors
We have $n$ samples to train ($x_1,\ldots,x_n$) and one to test ($x_0$)

each one has a label $y_1,\ldots,y_n$, and $y_0$

We measure the distance $d(x_0,x_i)$ for all $i\in \{1,\ldots,n\}$

We choose the $k$ elements closer to $x_0$, and we *vote* for the most frequent label. That is the predicted label $\hat{y_0}$


## Example with CMB1
```{r}
md5 <- read.csv("CMB1-2018-md5.tsv", sep="", stringsAsFactors=FALSE)
delivery <- read.delim("CMB1-2018-log.txt", header=FALSE,
                       stringsAsFactors=FALSE)
colnames(delivery) <- c("event", "md5", "ext", "datetime","email",
                        "filename", "mesg_id")
delivery$when <- strptime(delivery$datetime, 
                          format="%a, %e %b %Y %H:%M:%S %z")
combined <- merge(delivery, md5)[, c("number","event","when")]
combined$secs <- combined$when - min(combined$when)
```

## Example 

```{r message=FALSE}
library(reshape2)
pivot <- acast(combined, number~event, fun.aggregate = min,
               value.var = "secs")
delivery_delay <- sweep(pivot, 2, apply(pivot, 2, min))
delivered <- data.frame(Numara=rownames(delivery_delay),
                        1 - is.infinite(delivery_delay))
delivery_delay <- data.frame(Numara=rownames(delivery_delay),
                             1 - is.infinite(delivery_delay))
```

## Example 
```{r message=FALSE}
# midterm <- read.csv("CMB1-2018 - Midterm.csv", skip = 2, stringsAsFactors = FALSE)
library(readr)
midterm <- read_csv("CMB1-2018 - Midterm.csv", skip = 2)[1:63,]
colnames(midterm)[19] = "avg"
```

## Example 
```{r}
attendance <- read_csv("CMB1-2018 - Summary.csv", 
    col_types = cols(`%` = col_skip(), Sıra = col_skip()),
    skip = 2, n_max=70)[,-(15:28)]
```

## Example 
```{r}
m <- merge(midterm[, c("Numara", "avg")], attendance[,-(1:2)])
m <- merge(m, delivered, all.x = TRUE)
m[is.na(m)]  <- 0
```
